---
title: "Check Point #03 - Dataset Exploration"
author: "Justin Jialang Liu"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: inline
output:
  pdf_document:
    toc: true
    toc_depth: 1
    number_sections: true
    latex_engine: xelatex
urlcolor: blue
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(readr)

```

# Introduction

Sections from previous reports can be re-used if relevant. Any section of the report can be revised as the project develops. The goal3of this Check Point is to further data set exploration, and gain experience production publication-quality reports with R markdown. It is also an opportunity to prepare for future reports focused on statistical analyses.

# Research Question and Background

## Research Question

In the mWEL project’s target population of English-speaking adults, does perceived stress/distress differ by age group (\<25, 25–34, 35–44, 45–54, 55+)?

Operationally:

```         
•   A (predictor): age_group (categorical)

•   B (outcome): stress_index (composite of prodis1–prodis6, higher = more perceived distress)
```

## Scientific Plausibility and Context

Age is a key variable when evaluating whether a digital mental-health toolkit can improve teachers’ well-being because it plausibly shapes both baseline need and the pathway through which the intervention works. Teachers experience high levels of stress and burnout, and teacher burnout profiles are well documented in the literature, making it reasonable to expect meaningful heterogeneity in distress across demographic strata such as age.¹ At the same time, digital toolkits rely on users’ willingness and ability to engage with technology: in education contexts, teachers’ digital competence and technology acceptance are important determinants of intention to use and actual uptake of digital tools, which may vary with career stage and prior technology exposure—factors that often correlate with age.² Finally, evidence from digital mental health intervention research indicates that older adults can face distinct usability, trust, and accessibility barriers, implying that the same toolkit could show different engagement patterns and benefits across age groups.³ Taken together, age is scientifically important not only as a predictor of teachers’ stress levels but also as a potential effect modifier that can help identify which teacher subgroups are most likely to adopt the toolkit and experience measurable improvements.

References

1.  Agyapong B, Obuobi-Donkor G, Burback L, Wei Y. Stress, Burnout, Anxiety and Depression among Teachers: A Scoping Review. Int J Environ Res Public Health. 2022;19(17):10706. <doi:10.3390/ijerph191710706>.

2.  Antonietti C, Colombo B, Cremonesi P. Can teachers’ digital competence influence technology acceptance and use intention in vocational education and training? Comput Human Behav. 2022;132:107266. <doi:10.1016/j.chb.2022.107266>.

3.  Yin R, et al. The views and experiences of older adults regarding digital mental health interventions: a systematic review of qualitative evidence. Lancet Healthy Longev. 2024.

# Data Preparation

For the following sections, we require that at least one of the considered variables (B) to be categorical for summary and visualization purpose. If the research question focuses on continuous variables, one of the variables can be discretized.

## Loading Data

```{r,results='hide'}

library(haven)
library(dplyr)
library(tidyverse)
library(psych)


mywd <- "/Users/justinl/Desktop/Acadamic/Cornell/25-26 Spring/6270 R/Dataset Exploration"
setwd(mywd)

df <- read.csv("mWEL_R21_Data_1.csv")
glimpse(df)

```

## Creating avaerage score for Perceived Stress

```{r,results='hide'}

library(dplyr)

# We will create an average score for Perceived Stress
analysis_data <- df %>%
  mutate(
    # Composite for Perceived Distress (prodis1 to prodis6)
    stress_index = rowMeans(select(., prodis1:prodis6), na.rm = TRUE),
    
    # Ensure Age is numeric
    age_numeric = as.numeric(age)
  ) %>%
  # Filter out rows with missing values in our primary variables
  filter(!is.na(stress_index), !is.na(age_numeric))

```

## Categorization of Age

```{r,results='hide'}

library(dplyr)

analysis_data <- analysis_data %>%
  mutate(
    age_group = case_when(
      age_numeric < 25 ~ "<25",
      age_numeric >= 25 & age_numeric <= 34 ~ "25-34",
      age_numeric >= 35 & age_numeric <= 44 ~ "35-44",
      age_numeric >= 45 & age_numeric <= 54 ~ "45-54",
      age_numeric >= 55 ~ "55+",
      TRUE ~ NA_character_
    )
  )

```

## Quick look

```{r}
# Quick look at how these three variables relate linearly
cor_matrix <- analysis_data %>%
  select(age_numeric,stress_index) %>%
  cor(use = "complete.obs")
print(cor_matrix)

# Check how many people in each age groups
table(analysis_data$age_group, useNA = "ifany")

```

# Group Summary Statistics

```{r}

library(gtsummary)
library(dplyr)
library(kableExtra)

tab_age <- analysis_data %>%
  select(age_group, stress_index) %>%
  tbl_summary(
    by = age_group,
    type = list(stress_index ~ "continuous2"),
    statistic = list(
      all_continuous2() ~ c(
        "{mean} ({sd})",
        "{median} [{p25}, {p75}]",
        "{min}, {max}"
      )
    ),
    digits = all_continuous2() ~ 2
  ) %>%
  modify_header(label ~ "**Characteristic**") %>%
  modify_table_body(~ .x %>%
    mutate(
      label = case_when(
        label == "stress_index" & row_type == "label" ~ "Stress index",
        row_type == "level" & label == "Mean (SD)" ~ "  Mean (SD)",
        row_type == "level" & label == "Median (IQR)" ~ "  Median [Q1, Q3]",
        row_type == "level" & label == "Range" ~ "  Min, Max",
        TRUE ~ label
      )
    )
  ) %>%
  as_kable_extra(booktabs = TRUE) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 9
  )

tab_age

```

Across the five age categories, the descriptive statistics suggest only a weak association between age and average perceived distress (stress_index). Mean stress_index values are fairly similar across groups (roughly 2.5–2.7), with a slight uptick in the oldest group (55+) and the lowest mean in the 25–34 group, but the differences are small and the distributions overlap substantially (similar medians and overlapping IQRs across categories). This pattern is consistent with the near-zero correlation between continuous age and stress_index (r \~ 0.09), indicating that age alone is not strongly related to stress levels in this sample. Because the youngest (\<25) and oldest (55+) groups have smaller sample sizes, their estimates may be less stable and more sensitive to outliers.

# Comparative Visualizations

```{r}

library(ggplot2)
library(showtext)
showtext_auto()

pal <- c(
  "<25"   = "#4E79A7",
  "25-34" = "#59A14F",
  "35-44" = "#F28E2B",
  "45-54" = "#E15759",
  "55+"   = "#B07AA1"
)

p_violin <- ggplot(analysis_data, aes(x = age_group, y = stress_index)) +
  geom_violin(aes(fill = age_group), trim = FALSE, alpha = 0.80, linewidth = 0.25) +
  geom_boxplot(width = 0.16, outlier.shape = NA, alpha = 0.25, linewidth = 0.25) +
  # scatter points on top (jittered)
  geom_point(
    aes(color = "black"),
    position = position_jitter(width = 0.12, height = 0),
    alpha = 0.5,
    size = 1.2
  ) +
  stat_summary(fun = mean, geom = "point", color = "black", size = 1.7) +
  scale_fill_manual(values = pal) +
  scale_color_manual(values = pal) +
  guides(fill = "none", color = "none") +
  labs(
    x = "Age group (years)",
    y = "Perceived distress (stress_index)",
    title = "Perceived distress by age group"
  ) +
  theme_classic(base_family = "Arial", base_size = 8) +
  theme(
    plot.title = element_text(size = 9),
    axis.title = element_text(size = 8),
    axis.text  = element_text(size = 7)
  )


p_violin

```

The violin plot suggests only modest differences in perceived distress across age groups in the mWEL English-speaking adult sample. The central tendency (median and mean) is fairly similar for ages \<25 through 45–54, with heavy overlap in the interquartile ranges, implying that age alone is not strongly separating stress levels in this dataset. The 55+ group shows a slightly higher center (both mean/median appear a bit elevated) compared with the younger groups, but the distributions still overlap substantially, so any age gradient looks small rather than dramatic. The spread within each age group is wide relative to the between-group shifts, indicating substantial within-age heterogeneity in perceived distress.

# AI Use Disclosure Statement

This document was prepared with assistance from ChatGPT. I used ChatGPT to help debug R code for data cleaning (including recoding and age-group categorization), generating descriptive tables, and creating ggplot2 visualizations. All code was reviewed, tested, and modified by me to ensure it met the assignment requirements and worked correctly with the dataset.